---
title: "Pattern Matching with Regular Expression"
output: html_notebook
---
Matching the start or end of the string
rebus provides START and END shortcuts to specify regular expressions that match the start and end of the string. These are also known as anchors. You can try it out just by typing

START
You'll see the output <regex> ^. The <regex> denotes this is a special regex object and it has the value ^. ^ is the character used in the regular expression language to denote the start of a string.

The special operator provided by rebus, %R% allows you to compose complicated regular expressions from simple pieces. When you are reading rebus code, think of %R% as "then". For example, you could combine START with c,

START %R% "c"
to match the pattern "the start of string then a c", or in other words: strings that start with c. In rebus, if you want to match a specific character, or a specific sequence of characters, you simply specify them as a string, e.g. surround them with ".

```{r}
library(rebus)
library(stringr)

# Some strings to practice with
x <- c("cat", "coat", "scotland", "tic toc")

# Print END
END

# Run me
str_view(x, pattern = START %R% "c")

# Match the strings that start with "co" 
str_view(x, pattern = START %R% "co")

# Match the strings that end with "at"
str_view(x, pattern = "at" %R% END)

# Match the strings that is exactly "cat"
str_view(x, pattern = START %R% "cat" %R% END)
```

Matching any character
In a regular expression you can use a wildcard to match a single character, no matter what the character is. In rebus it is specified with ANY_CHAR. Try typing ANY_CHAR in the console. You should see that in the regular expression language this is specified by a dot, ..

For example, "c" %R% ANY_CHAR %R% "t" will look for patterns like "c_t" where the blank can be any character. Consider the strings: "cat", "coat", "scotland" and "tic toc". Where would the matches to "c" %R% ANY_CHAR %R% "t" be?

Test your intuition by running:

str_view(c("cat", "coat", "scotland", "tic toc"), 
  pattern = "c" %R% ANY_CHAR %R% "t")
Notice that ANY_CHAR will match a space character (c t in tic toc). It will also match numbers or punctuation symbols, but ANY_CHAR will only ever match one character, which is why we get no match in coat.

```{r}
x <- c("cat", "coat", "scotland", "tic toc")

# Match any character followed by a "t"
str_view(x, pattern = ANY_CHAR %R% "t")

# Match a "t" followed by any character
str_view(x, pattern = "t" %R% ANY_CHAR)

# Match two characters
str_view(x, pattern = ANY_CHAR %R% ANY_CHAR)

# Match a string with exactly three characters
str_view(x, pattern = START %R% ANY_CHAR %R% ANY_CHAR %R% ANY_CHAR %R% END)
```

Combining with stringr functions
You can pass a regular expression as the pattern argument to any stringr function that has the pattern argument. You can use str_detect() to get a logical vector for whether there was a match, str_subset() to return just the strings with matches, and str_count() to count the number of matches in each string.

As a reminder, compare the output of those three functions with our "c_t" pattern from the previous exercise:

x <- c("cat", "coat", "scotland", "tic toc")
pattern <- "c" %R% ANY_CHAR %R% "t"
str_detect(x, pattern)
str_subset(x, pattern)
str_count(x, pattern)
It now also makes sense to add str_extract() to your repertoire. It returns just the part of the string that matched the pattern:

str_extract(x, pattern)
You'll combine your regular expression skills with stringr to ask how often a q is followed by a u in boy names.

```{r}
library(babynames)
library(dplyr)
# Extracting vectors for boys' names
babynames_2014 <- filter(babynames, year == 2014)
boy_names = filter(babynames_2014, sex == "M")$name
# q followed by any character
pattern <- "q" %R% ANY_CHAR

# Test pattern 
str_view(c("Quentin", "Kaliq", "Jacques",  "Jacqes"), pattern)  

# Find names that have the pattern
names_with_q <- str_subset(boy_names, pattern = pattern)
length(names_with_q)

# Find part of name that matches pattern
part_with_q <- str_extract(boy_names, pattern = pattern)
table(part_with_q)

# Did any names have the pattern more than once?
count_of_q <- (str_count(boy_names, pattern = pattern))
table(count_of_q)
# Which babies got these names?
with_q <- str_detect(boy_names, pattern = pattern)

# What fraction of babies got these names?
mean(with_q)
```

Alternation
The rebus function or() allows us to specify a set of alternatives, which may be single characters or character strings, to be matched. Each alternative is passed as a separate argument.

For example, or("grey", "gray") allows us to detect either the American or British spelling:

x <- c("grey sky", "gray elephant")
str_view(x, pattern = or("grey", "gray"))
Since these two words only differ by one character you could equivalently specify this match with "gr" %R% or("e", "a") %R% "y", that is "a gr followed by, an e or an a, then a y".

```{r}
girl_names <- filter(babynames_2014, sex == "F")$name

# Match Jeffrey or Geoffrey
whole_names <- or("Jeffrey", "Geoffrey")
str_view(boy_names, pattern = whole_names, 
  match = TRUE)

# Match Jeffrey or Geoffrey, another way
common_ending <- or("Je","Geo") %R% "ffrey"
str_view(boy_names, pattern = common_ending, 
  match = TRUE)

# Match with alternate endings
by_parts <- or("Je", "Geo") %R% "ff" %R% or("ry", "ery", "rey", "erey")
str_view(boy_names, 
  pattern = by_parts, 
  match = TRUE)

# Match names that start with Cath or Kath
ckath <- or("C", "K") %R% "ath"
str_view(girl_names, pattern = ckath, match = TRUE)

```

Character classes
In regular expressions a character class is a way of specifying "match one (and only one) of the following characters". In rebus you can specify the set of allowable characters using the function char_class().

This is another way you could specify an alternate spelling, for example, specifying "a gr followed by, either an a or e, followed by a y":

x <- c("grey sky", "gray elephant")
str_view(x, pattern = "gr" %R% char_class("ae") %R% "y")
A negated character class matches "any single character that isn't one of the following", and in rebus is specified with negated_char_class().

Unlike in other places in a regular expression you don't need to escape characters that might otherwise have a special meaning inside character classes. If you want to match . you can include . directly, e.g. char_class("."). Matching a - is a bit trickier. If you need to do it, just make sure it comes first in the character class.

```{r}
x <- c("grey sky", "gray elephant")
# Create character class containing vowels
vowels <- char_class("a", "e", "i", "o", "u", "A", "E", "I", "O", "U")

# Print vowels
vowels

# See vowels in x with str_view()
str_view(x, pattern = vowels)

# See vowels in x with str_view_all()
str_view_all(x, pattern = vowels)

# Number of vowels in boy_names
num_vowels <- str_count(boy_names, pattern = vowels)
mean(num_vowels)

# Proportion of vowels in boy_names
name_length <- str_length(boy_names)
mean(num_vowels/name_length)
```

Repetition
The rebus functions one_or_more(), zero_or_more() and optional() can be used to wrap parts of a regular expression to allow a pattern to match a variable number of times.

Take our vowels pattern from the last exercise. You can pass it to one_or_more() to create the pattern that matches "one or more vowels". Take a look with these interjections:

x <- c("ow", "ooh", "yeeeah!", "shh")
str_view(x, pattern = one_or_more(vowels))
You'll see we can match the single o in ow, the double o in ooh and the string of es followed by the a in yeeeah, but nothing in shh because there isn't a single vowel.

In contrast zero_or_more() will match even if there isn't an occurrence, try

str_view(x, pattern = zero_or_more(vowels))
Since both yeeeah and shh start without a vowel, they match "zero vowels", and since regular expressions are lazy, they look no further and return the start of the string as a match.

```{r}
# Vowels from last exercise
vowels <- char_class("aeiouAEIOU")

# Use `negated_char_class()` for everything but vowels
not_vowels <- negated_char_class("aeiouAEIOU")

# See names with only vowels
str_view(boy_names, 
  pattern = exactly(one_or_more(vowels)), 
  match = TRUE)

# See names with no vowels
str_view(boy_names, 
  pattern = exactly(one_or_more(not_vowels)), 
  match = TRUE)
```

Hunting for phone numbers
For your first task you are going to pull out the phone numbers from this vector of contact information:

contact
You'll assume the phone numbers you are looking for follow the American standard of a three digit area code, a three digit exchange and then a four digit number, but each part could be separated by spaces or various punctuation symbols.

```{r}
contact <- c("Call me at 555-555-0191", "123 Main St", "(555) 555 0191",  "Phone: 555.555.0191 Mobile: 555.555.0192")
# Take a look at ALL digits
str_view_all(contact, DGT)

# Create a three digit pattern and test
three_digits <- DGT %R% DGT %R% DGT
str_view_all(contact,
  pattern = three_digits)

# Create four digit pattern
four_digits <- three_digits %R% DGT

# Create a separator pattern and test
separator <- char_class("-.() ")
str_view_all(contact,
  pattern = separator)

# Create phone pattern
phone_pattern <- optional(OPEN_PAREN) %R% 
  three_digits %R% 
  zero_or_more(separator) %R% 
  three_digits %R% 
  zero_or_more(separator) %R%
  four_digits
        
# Test pattern           
str_view(contact, pattern = phone_pattern)

# Extract phone numbers
str_extract(contact, phone_pattern)

# Extract ALL phone numbers
str_extract_all(contact, phone_pattern)
```

Extracting age and gender from accident narratives
Recall from the video, you want to parse out age and gender from accident narratives. For example, this narrative

19YOM-SHOULDER STRAIN-WAS TACKLED WHILE PLAYING FOOTBALL W/ FRIENDS 
describes male of age 19, and this one

TRIPPED OVER CAT AND LANDED ON HARDWOOD FLOOR. LACERATION ELBOW, LEFT. 33 YOF*
a female of age 33.

You are generally looking for a pattern with a number, something to indicate the units, e.g. YO or YR for years old, or MO for months old, and a character that identifies the gender.

In this exercise you'll build up a pattern to pull out the part of the narrative that has the age and gender information. Then, in the next exercise you'll parse out the age and gender into separate variables.

```{r}
narratives <- c("19YOM-SHOULDER STRAIN-WAS TACKLED WHILE PLAYING FOOTBALL W/ FRIENDS ", "31 YOF FELL FROM TOILET HITITNG HEAD SUSTAINING A CHI ",  "ANKLE STR. 82 YOM STRAINED ANKLE GETTING OUT OF BED ", "TRIPPED OVER CAT AND LANDED ON HARDWOOD FLOOR. LACERATION ELBOW, LEFT. 33 YOF*", "10YOM CUT THUMB ON METAL TRASH CAN DX AVULSION OF SKIN OF THUMB ", "53 YO F TRIPPED ON CARPET AT HOME. DX HIP CONTUSION ", "13 MOF TRYING TO STAND UP HOLDING ONTO BED FELL AND HIT FOREHEAD ON RADIATOR DX LACERATION", "14YR M PLAYING FOOTBALL; DX KNEE SPRAIN ", "55YOM RIDER OF A BICYCLE AND FELL OFF SUSTAINED A CONTUSION TO KNEE ", "5 YOM ROLLING ON FLOOR DOING A SOMERSAULT AND SUSTAINED A CERVICAL STRA IN")

# Look for two digits
str_view(narratives, DGT %R% DGT)

# Pattern to match one or two digits
age <- optional(DGT) %R% DGT
str_view(narratives, 
  pattern = age)

# Pattern to match units 
unit <- optional(SPC) %R% or("YO", "YR", "MO")

# Test pattern with age then units
str_view(narratives, 
  pattern = age %R% unit)

# Pattern to match gender
gender <- optional(SPC) %R% or("M", "F")

# Test pattern with age then units then gender
str_view(narratives, 
  pattern = age %R% unit %R% gender)

# Extract age_gender, take a look
age_gender <- str_extract(narratives,  age %R% unit %R% gender)
age_gender

```

Parsing age and gender into pieces
To finish up, you need to pull out the individual pieces and tidy them into usable variables.

There are a few ways you could get at one piece: you could extract out the piece you need, you could replace everything that isn't the piece you need with "", or you could try to split into the pieces you need. You'll try a few of these in this exercise and you'll see yet another way in the next chapter.

One benefit of building up your pattern in pieces is you already have patterns for each part that you can reuse now.

```{r}
# age_gender, age, gender, unit are pre-defined
ls.str()

# Extract age and make numeric
ages_numeric <- as.numeric(str_extract(age_gender, pattern = age))

# Replace age and units with ""
genders <- str_replace(age_gender, 
  pattern = age %R% unit, 
  replacement = "")

# Replace extra spaces
genders_clean <- str_replace_all(genders, 
  pattern = optional(SPC), 
  replacement = "")

# Extract units 
time_units <- str_extract(age_gender, pattern = unit)

# Extract first word character
time_units_clean <- str_extract(time_units, pattern = WRD)

# Turn ages in months to years
ages_years <- ifelse(time_units_clean == "Y", ages_numeric, ages_numeric/12)
```

Capturing parts of a pattern
In rebus, to denote a part of a regular expression you want to capture, you surround it with the function capture(). For example, a simple pattern to match an email address might be,

email <- one_or_more(WRD) %R% 
  "@" %R% one_or_more(WRD) %R% 
  DOT %R% one_or_more(WRD)
str_view("(wolverine@xmen.com)", pattern = email)  
If you want to capture the part before the @, you simply wrap that part of the regular expression in capture():

email <- capture(one_or_more(WRD)) %R% 
  "@" %R% one_or_more(WRD) %R% 
  DOT %R% one_or_more(WRD)
str_view("(wolverine@xmen.com)", pattern = email)  
The part of the string that matches hasn't changed, but if we pull out the match with str_match() we get access to the captured piece:

str_match("(wolverine@xmen.com)", pattern =  email)  
You'll explore this behavior with some more super hero email addresses.

```{r}
hero_contacts <- c("(wolverine@xmen.com)", "wonderwoman@justiceleague.org", "thor@avengers.com")
# Capture part between @ and . and after .
email <- capture(one_or_more(WRD)) %R% 
  "@" %R% capture(one_or_more(WRD)) %R% 
  DOT %R% capture(one_or_more(WRD))

# Check match hasn't changed
str_view(hero_contacts, pattern = email)

# Pull out match and captures
email_parts <- str_match(hero_contacts, pattern = email)

# Print email_parts
email_parts

# Save host
host <- email_parts[,3]
host
```

Pulling out parts of a phone number
You can now go back to the phone number example from the previous chapter. You developed a pattern to extract the parts of string that looked like phone numbers, and now you have the skills to pull out the pieces of the number. Let's see if you can put your skills together to output the first phone number in each string in a common format.

```{r}
# View text containing phone numbers
contact

# Add capture() to get digit parts
phone_pattern <- capture(three_digits) %R% zero_or_more(separator) %R% 
           capture(three_digits) %R% zero_or_more(separator) %R%
           capture(four_digits)
           
# Pull out the parts with str_match()
phone_numbers <- str_match(contact, pattern = phone_pattern)

# Put them back together
str_c(
  "(",
  phone_numbers[,2],
  ")",
  phone_numbers[,3],
  "-",
  phone_numbers[,4])
```

Extracting age and gender again
You can now also take another look at your pattern for pulling out age and gender from the injury narratives. In the last chapter you took two steps to extract the pattern and then pull it apart. Take a go at doing it in one step!

```{r}
# narratives has been pre-defined
narratives

# Add capture() to get age, unit and sex
pattern <- capture(optional(DGT) %R% DGT) %R%  
  optional(SPC) %R% capture(or("YO", "YR", "MO")) %R%
  optional(SPC) %R% capture(or("M", "F"))

# Pull out from narratives
str_match(narratives, pattern)

# Edit to capture just Y and M in units
pattern2 <- capture(optional(DGT) %R% DGT) %R%  
  optional(SPC) %R% capture(or("Y", "M")) %R% optional(or("O","R")) %R%
  optional(SPC) %R% capture(or("M", "F"))

# Check pattern
str_view(narratives, pattern2)

# Pull out pieces
str_match(narratives, pattern2)
```

```{r}
# See names with three repeated letters
repeated_three_times <- capture(LOWER) %R% REF1 %R% REF1
str_view(boy_names, 
  pattern = repeated_three_times, 
  match = TRUE)

# See names with a pair of repeated letters
pair_of_repeated <- capture(LOWER %R% LOWER) %R% REF1
str_view(boy_names, 
  pattern = pair_of_repeated, 
  match = TRUE)

# See names with a pair that reverses
pair_that_reverses <- capture(LOWER) %R% capture(LOWER) %R% REF2 %R% REF1
str_view(boy_names, 
  pattern = pair_that_reverses, 
  match = TRUE)

# See four letter palindrome names
four_letter_palindrome <- exactly(capture(LOWER) %R% capture(LOWER) %R% 
      REF2 %R% REF1)
str_view(boy_names, 
  pattern = four_letter_palindrome, 
  match = TRUE)

# See six letter palindrome names
six_letter_palindrome <- exactly(
    capture(LOWER) %R% capture(LOWER) %R% capture(LOWER) %R% 
      REF3 %R% REF2 %R% REF1)
str_view(boy_names, 
  pattern = six_letter_palindrome, 
  match = TRUE)
```

Replacing with regular expressions
Now, you've mastered matching with backreferences, you'll build up to replacing with backreferences, but first let's review str_replace() now that you've got regular expressions under your belt.

Remember str_replace() takes three arguments, string a vector of strings to do the replacements in, pattern that identifies the parts of strings to replace and replacement the thing to use as a replacement.

replacement can be a vector, the same length as string, each element specifies the replacement to be used in each string. Let's practice by anonymizing some of the contact objects you've seen so far.

```{r}
# View text containing phone numbers
contact

# Replace digits with "X"
str_replace(contact, DGT, "X")

# Replace all digits with "X"
str_replace_all(contact, DGT, "X")

# Replace all digits with different symbol
str_replace_all(contact, DGT, c("X", ".", "*", "_"))
```

Replacing with backreferences
The replacement argument to str_replace() can also include backreferences. This works just like specifying patterns with backreferences, except the capture happens in the pattern argument, and the backreference is used in the replacement argument.

x <- c("hello", "sweet", "kitten")
str_replace(x, capture(ANY_CHAR), str_c(REF1, REF1))
capture(ANY_CHAR) will match the first character no matter what it is. Then the replacement str_c(REF1, REF1) combines the captured character with itself, in effect doubling the first letter of each string.

You are going to use this to create some alternative, more lively accident narratives.

The strategy you'll take is to match words ending in "ING" then replace them with an adverb followed by the original word.

```{r}
adverbs = readRDS("adverbs.rds")
adverbs

# Build pattern to match words ending in "ING"
pattern <- one_or_more(WRD) %R% "ING"
str_view(narratives, pattern)

# Test replacement
str_replace(narratives, capture(pattern), str_c("CARELESSLY", REF1, sep = " "))

# One adverb per narrative
adverbs_10 <- sample(adverbs, 10)

# Replace "***ing" with "adverb ***ing"
str_replace(narratives, capture(pattern), str_c(adverbs_10, REF1, sep = " "))
```

Matching a specific code point or code groups
Things can get tricky when some characters can be specified two ways, for example ??, an e with a grave accent, can be specified either with the single code point \u00e8 or the combination of a \u0065 and a combining grave accent \u0300. They look the same:

x <- c("\u00e8", "\u0065\u0300")
writeLines(x)
But, specifying the single code point only matches that version:

str_view(x, "\u00e8")
The stringi package that stringr is built on contains functions for converting between the two forms. stri_trans_nfc() composes characters with combining accents into a single character. stri_trans_nfd() decomposes character with accents into separate letter and accent characters. You can see how the characters differ by looking at the hexadecimal codes.

as.hexmode(utf8ToInt(stri_trans_nfd("\u00e8")))
as.hexmode(utf8ToInt(stri_trans_nfc("\u0065\u0300")))
In Unicode, an accent is known as a diacritic Unicode Property, and you can match it using the rebus value UP_DIACRITIC.

Vietnamese makes heavy use of diacritics to denote the tones in its words. In this exercise, you'll manipulate the diacritics in the names of Vietnamese rulers.

```{r}
library(stringi)
# Names with builtin accents
(tay_son_builtin <- c(
  "Nguy\u1ec5n Nh\u1ea1c", 
  "Nguy\u1ec5n Hu\u1ec7",
  "Nguy\u1ec5n Quang To\u1ea3n"
))

# Convert to separate accents
tay_son_separate <- stri_trans_nfd(tay_son_builtin)

# Verify that the string prints the same
tay_son_separate

# Match all accents
str_view_all(tay_son_separate, pattern = UP_DIACRITIC)
```

Matching a single grapheme
A related problem is matching a single character. You've used ANY_CHAR to do this up until now, but it will only match a character represented by a single code point. Take these three names:

x <- c("Adele", "Ad\u00e8le", "Ad\u0065\u0300le")
writeLines(x)
They look the similar, but this regular expression only matches two of them:

str_view(x, "Ad" %R% ANY_CHAR %R% "le")
because in the third name ?? is represented by two code points. The unicode standard has a concept of a grapheme that represents a display character, but may be composed of many code points. To match any grapheme you can use GRAPHEME.

str_view(x, "Ad" %R% GRAPHEME %R% "le")

```{r}
# tay_son_separate has been pre-defined
tay_son_separate

# View all the characters in tay_son_separate
str_view_all(tay_son_separate, pattern = ANY_CHAR)

# View all the graphemes in tay_son_separate
str_view_all(tay_son_separate, pattern = GRAPHEME)

# Combine the diacritics with their letters
tay_son_builtin <- stri_trans_nfc(tay_son_separate)

# View all the graphemes in tay_son_builtin
str_view_all(tay_son_builtin, patter = GRAPHEME)
```

